Hierarchical Dispatching and Overriding

### 2017-06-18 Aplas17/paper.tex
Add text for implementaion.
separate Aux defs out to a section
fix Aux
remove example
grammar check Sec1,Sec2

\newcommand\bruno[1]{\authornote{bruno}{red}{#1}}
\newcommand\yanlin[1]{\authornote{yanlin}{purple}{#1}}
\newcommand\marco[1]{\authornote{marco}{blue}{#1}}
\newcommand\haoyuan[1]{\authornote{haoyuan}{cyan}{#1}}

% \newcommand\bruno[1]{}
% \newcommand\yanlin[1]{}
% \newcommand\marco[1]{}
% \newcommand\haoyuan[1]{}

@@

\begin{abstract}
Multiple inheritance is a valuable feature for Object-Oriented
Programming. However it is also tricky to get right, as illustrated by
the extensive literature on the topic. One of the most promising 
approaches to multiple inheritance are \emph{traits}. Traits offer a
restricted model of multiple inheritance that is easy to reason and 
has many elegant properties. Traits have good support for method 
conflicts which arise from \emph{diamond inheritance}:
conflicts that arise from method implementations sharing a common 
ancestor. However the mechanisms of traits are inadequate to deal 
with \emph{unintentional method conflicts}: conflicts which 
arise from two unrelated methods that happen to share the same name
and signature.

This paper presents a new model \name{} 
(short for \emph{\textbf{F}eatherweight \textbf{H}ierarchical \textbf{J}ava}) 
that deals with unintentional
method conflicts. In our new model conflicting methods arising
from unrelated methods can coexist in the same component. Moreover they
can be \emph{independently overridden}. What ensures unambiguity 
is the use of information about the class hierarchy. In turn
hierarchical information is also employed in method dispatching, which 
uses a combination of static and dynamic type information to choose 
the implementation of a method at run-time. We give illustrative 
examples of our language and features and formalize \name{} as a minimal 
Featherweight-Java style calculus.

@@

\input{sections/Formalization.tex}
% \input{sections/TypeSafety.tex}

>>>>

% \newcommand\bruno[1]{\authornote{bruno}{red}{#1}}
% \newcommand\yanlin[1]{\authornote{yanlin}{purple}{#1}}
% \newcommand\marco[1]{\authornote{marco}{blue}{#1}}
% \newcommand\haoyuan[1]{\authornote{haoyuan}{orange}{#1}}

\newcommand\bruno[1]{}
\newcommand\yanlin[1]{}
\newcommand\marco[1]{}
\newcommand\haoyuan[1]{}

@@

\begin{abstract}
Multiple inheritance is a valuable feature for Object-Oriented
Programming. However, it is also tricky to get right, as illustrated by
the extensive literature on the topic. One of the most promising 
approaches to multiple inheritance is the \emph{trait} model. Traits offer a
restricted model of multiple inheritance that is easy to reason and 
has many elegant properties. Traits have good support for method 
conflicts which arise from \emph{diamond inheritance}:
conflicts that arise from method implementations sharing a common 
ancestor. However, the mechanisms of traits are inadequate to deal 
with \emph{unintentional method conflicts}: conflicts which 
arise from two unrelated methods that happen to share the same name
and signature.

This paper presents a new model \name{} 
(short for \emph{\textbf{F}eatherweight \textbf{H}ierarchical \textbf{J}ava}) 
that deals with unintentional
method conflicts. In our new model, conflicting methods arising
from unrelated methods can coexist in the same component. Moreover, they
can be \emph{independently overridden}. What ensures unambiguity 
is the use of information about the class hierarchy. In turn, 
hierarchical information is also employed in method dispatching, which 
uses a combination of static and dynamic type information to choose 
the implementation of a method at run-time. We give illustrative 
examples of our language and features and formalize \name{} as a minimal 
Featherweight-Java style calculus.

@@

\input{sections/Formalization.tex}
\input{sections/Auxiliary.tex}
% \input{sections/TypeSafety.tex}

### 2017-06-18 Aplas17/sections/Introduction.tex
grammar check Sec1,Sec2

\section{Introduction}
Inheritance in Object-Oriented Programming (OOP) offers a mechanism
for code reuse. However many OOP languages are restricted to single
inheritance, which is less expressive and flexible than multiple
inheritance. Nevertheless different flavours of multiple inheritance
have been adopted in some popular OOP languages. C++ has had 
multiple inheritance from the start. Scala adapts ideas from traits~\cite{scharli03traits} 
and mixins~\cite{bracha90mixin} to offer a disciplined form of multiple inheritance. Java 8 
offers a simple variant of traits with disguised of interfaces with default methods~\cite{goetz12fdefenders}.

@@

In contrast with the diamond inheritance, a second case of ambiguity
is \textit{unintentional method conflicts}~\cite{scharli03traits}. That is conflicting 
methods that do not actually refer to the same feature. 
In a nominal system, methods can be designed for different
functionality, but they just happen to have same names (and
signatures).
A simple example of this situation are two \lstinline{draw} methods that
are inherited from a deck of cards and a drawable widget. 
In such context the two \lstinline{draw} methods have very different meanings, 
but they happen to share the same name.
%%This issue was proposed by the trait paper, so-called
When inheritance is used to compose these methods, a compilation 
error happens due to conflicts. However, unlike the diamond problem,
the conflicting methods have very different meanings and do not share a
common parent. We call such a case ``\textit{triangle inheritance}''.
\haoyuan{Fine to have a graph with diamond and triangle shapes if we got space.}

@@

Some other workarounds or approaches include delegation, and
renaming/exclusion in the trait model. Yet they still have various
drawbacks as we will discuss in Section~\ref{sec:overview}. Closest to our work
are mechanisms available in C++ and C\# that allow for two
unintentionally conflicting methods to coexist in a class. Among them
C++ is a representative that accepts the triangle inheritance and
resolves the ambiguity by \emph{static dispatching}. However, C++ has
limited support for virtual methods with unintentional conflicts, and
it will often throw errors when composing them. This is again
unsatisfactory because virtual methods are pervasive in OOP and used 
for code reuse and extensibility. A problem with the C++ approach is
that programmers can only use either static or dynamic dispatching separately, but dealing
with unintentional method conflicts seems to require a combination of both.

@@

This paper proposes \textit{hierarchical dispatch}: a novel approach
to method dispatching, which combines static and dynamic
information. Using hierarchical dispatch, the method binder will look
at both the \emph{static type} and the \emph{dynamic type} of the
receiver during runtime. When there are multiple branches that cause
unintentional conflicts, the static type can specify one branch among
them for unambiguity, and the dynamic type helps to find the most
specific implementation. In that case both unambiguity and
extensibility are preserved. To present this idea, we introduce a
formalized model \MIM{} in Section~\ref{sec:formalization} based on
Featherweight Java~\cite{Igarashi01FJ}, together with theorems and
proofs for type soundness. In the model we also propose
\textit{hierarchical overriding}, where method overriding can be applied
only to one branch of the class hierarchy. Our model can be viewed as
a generalization of the trait model by providing additional support to
the triangle inheritance.

>>>>

\section{Introduction}
Inheritance in Object-Oriented Programming (OOP) offers a mechanism
for code reuse. However many OOP languages are restricted to single
inheritance, which is less expressive and flexible than multiple
inheritance. Nevertheless, different flavours of multiple inheritance
have been adopted in some popular OOP languages. C++ has had 
multiple inheritance from the start. Scala adapts ideas from traits~\cite{scharli03traits} 
and mixins~\cite{bracha90mixin} to offer a disciplined form of multiple inheritance. Java 8 
offers a simple variant of traits with disguised of interfaces with default methods~\cite{goetz12fdefenders}.

@@

In contrast with the diamond inheritance, the second case of ambiguity
is \textit{unintentional method conflicts}~\cite{scharli03traits}. That is conflicting 
methods that do not actually refer to the same feature. 
In a nominal system, methods can be designed for different
functionality, but they just happen to have same names (and
signatures).
A simple example of this situation is two \lstinline{draw} methods that
are inherited from a deck of cards and a drawable widget. 
In such context, the two \lstinline{draw} methods have very different meanings, 
but they happen to share the same name.
%%This issue was proposed by the trait paper, so-called
When inheritance is used to compose these methods, a compilation 
error happens due to conflicts. However, unlike the diamond problem,
the conflicting methods have very different meanings and do not share a
common parent. We call such a case ``\textit{triangle inheritance}''.
\haoyuan{Fine to have a graph with diamond and triangle shapes if we got space.}

@@

Some other workarounds or approaches include delegation, and
renaming/exclusion in the trait model. Yet they still have various
drawbacks as we will discuss in Section~\ref{sec:overview}. Closest to our work
are mechanisms available in C++ and C\# that allow for two
unintentionally conflicting methods to coexist in a class. Among them, 
C++ is a representative that accepts the triangle inheritance and
resolves the ambiguity by \emph{static dispatching}. However, C++ has
limited support for virtual methods with unintentional conflicts, and
it will often throw errors when composing them. This is again
unsatisfactory because virtual methods are pervasive in OOP and used 
for code reuse and extensibility. A problem with the C++ approach is
that programmers can only use either static or dynamic dispatching separately, but dealing
with unintentional method conflicts seems to require a combination of both.

@@

This paper proposes \textit{hierarchical dispatch}: a novel approach
to method dispatching, which combines static and dynamic
information. Using hierarchical dispatch, the method binder will look
at both the \emph{static type} and the \emph{dynamic type} of the
receiver during runtime. When there are multiple branches that cause
unintentional conflicts, the static type can specify one branch among
them for unambiguity, and the dynamic type helps to find the most
specific implementation. In that case, both unambiguity and
extensibility are preserved. To present this idea, we introduce a
formalized model \MIM{} in Section~\ref{sec:formalization} based on
Featherweight Java~\cite{Igarashi01FJ}, together with theorems and
proofs for type soundness. In the model we also propose
\textit{hierarchical overriding}, where method overriding can be applied
only to one branch of the class hierarchy. Our model can be viewed as
a generalization of the trait model by providing additional support to
the triangle inheritance.

### 2017-06-18 Aplas17/sections/Overview.tex
grammar check Sec1,Sec2

\section{A Running Example: DrawableDeck}~\label{sec:overview}
This section illustrates the features of our model for
resolving unintentional method conflicts. As mentioned before, such a
case arises when two inherited methods happen to have the same
signature, but with different semantics and functionality. This
situation is troublesome to programmers that use multiple
inheritance. Below we illustrate with a running example called
\lstinline|DrawableDeck|, which models a drawable deck of cards. 
Note that we use Java-like syntax
throughout the paper, and all types are defined with the keyword
``\lstinline|interface|''. The concept is closely related to Java 8
interfaces with default methods~\cite{bono14} and traits. For simplicity, an interface in our model has
the following characteristics:
\begin{itemize}
  \item It allows multiple inheritance.
  \item Every method represents a behaviour and requires a body for its implementation (like Java 8 default methods). There are no abstract methods.
  \item It can be directly instantiated by the \lstinline|new| keyword.
\end{itemize}
Furthermore for simplicity, our examples and formalization does not deal with state at
this stage. %%, but that should not block our discussions below.

\subsection{Problem 1: Unintentional Method Conflicts}
Suppose that two components \lstinline|Deck| and \lstinline|Drawable| 
have been developed in a system.  \lstinline|Deck| represents a deck
of cards, and defines \lstinline|draw| for drawing a card from the
deck.  \lstinline|Drawable| is an interface for graphics that
can be drawn, and also includes a method called \lstinline|draw| for
visual display. For illustration, the implementation of
\lstinline|draw| in \lstinline|Drawable| only creates a blank canvas
on the screen, and later some instances (\lstinline|Circle|,
\lstinline|Square|, or others) that extend \lstinline|Drawable| may
define their own \lstinline|drawObject| methods and invoke
\lstinline|draw|.

@@

Now one may quickly comes up with a workaround, which is creating a new \lstinline|draw| in \lstinline|DrawableDeck| to
explicitly overrides the old ones. However, the two conflicting methods have totally different functionalities, merging
them into one does not make any sense. This non-solution will hide the old methods, and break independent extensibility. The
feature we expect is called ``triangle inheritance'', that is to say, when the conflicting methods serve for different behaviours,
we can allow them to coexist in the extending interface in a type-safe way.\\

@@

In the following text we will see that C++ is still unsatisfactory, as it introduces another problem on code reuse, which motivates
our work.

@@

\begin{figure*}[t]
  \centering
  \includegraphics[height=2cm]{pics/DrawableLoggingDeck.pdf}
  \caption{UML graph for \lstinline|DrawableLoggingDeck|.}\label{fig:drawableloggingdeck}
\end{figure*}
If we try to keep the triangle inheritance, but apply dynamic dispatch, calling \lstinline|shuffleAndDraw| would trigger ambiguity. 
In principle we want \lstinline|shuffleAndDraw| to invoke \lstinline|LoggingDeck.draw|, because they belong to the same branch, 
nonetheless, choosing either static dispatch
or dynamic dispatch cannot ensure both the triangle inheritance and code reuse. As far as we know, C++ does not have a solution
to it. Therefore we need to find another algorithm for method resolution.\\

@@

On the other hand, \lstinline|e.m()| where \lstinline|e| has static type \lstinline|I|, simply behaves the same as \lstinline|e.I::m()| in our model. Such a dispatch algorithm makes use of both the static type and the dynamic type of the receiver, so it is like a combination of static and dynamic dispatch. Intuitively, the static type specifies one branch to avoid ambiguity, and the dynamic type finds the latest version on that branch. It may still introduce ambiguity when there are multiple paths from the static type to the dynamic type, and those paths cause conflicts. We disallow this kind of conflict to ensure unambiguity. That is to say, we do not allow two methods to override a same base method when they cause conflicts. This is natural, as they are just two versions of the same operation, hence it is no longer an ``unintentional'' conflict, but the diamond problem.

>>>>

\section{A Running Example: DrawableDeck}~\label{sec:overview}
This section illustrates the features of our model for
resolving unintentional method conflicts. As mentioned before, such a
case arises when two inherited methods happen to have the same
signature, but with different semantics and functionality. This
situation is troublesome to programmers that use multiple
inheritance. Below we illustrate with a running example called
\lstinline|DrawableDeck|, which models a drawable deck of cards. 
Note that we use Java-like syntax
throughout the paper, and all types are defined with the keyword
``\lstinline|interface|''. The concept is closely related to Java 8
interfaces with default methods~\cite{bono14} and traits. For simplicity, an interface in our model has
the following characteristics:
\begin{itemize}
  \item It allows multiple inheritance.
  \item Every method represents a behaviour and requires a body for its implementation (like Java 8 default methods). There are no abstract methods.
  \item It can be directly instantiated by the \lstinline|new| keyword.
\end{itemize}
Furthermore for simplicity, our examples and formalization do not deal with state at
this stage. %%, but that should not block our discussions below.

\subsection{Problem 1: Unintentional Method Conflicts}
Suppose that two components \lstinline|Deck| and \lstinline|Drawable| 
have been developed in a system.  \lstinline|Deck| represents a deck
of cards and defines \lstinline|draw| for drawing a card from the
deck.  \lstinline|Drawable| is an interface for graphics that
can be drawn, and also includes a method called \lstinline|draw| for
visual display. For illustration, the implementation of
\lstinline|draw| in \lstinline|Drawable| only creates a blank canvas
on the screen, and later some instances (\lstinline|Circle|,
\lstinline|Square|, or others) that extend \lstinline|Drawable| may
define their own \lstinline|drawObject| methods and invoke
\lstinline|draw|.

@@

Now one may quickly come up with a workaround, which is creating a new \lstinline|draw| in \lstinline|DrawableDeck| to
explicitly overrides the old ones. However, the two conflicting methods have totally different functionalities, merging
them into one does not make any sense. This non-solution will hide the old methods, and break independent extensibility. The
feature we expect is called ``triangle inheritance'', that is to say, when the conflicting methods serve for different behaviours,
we can allow them to coexist in the extending interface in a type-safe way.\\

@@

In the following text, we will see that C++ is still unsatisfactory, as it introduces another problem on code reuse, which motivates
our work.

@@

\begin{figure*}[t]
  \centering
  \includegraphics[height=2cm]{pics/DrawableLoggingDeck.pdf}
  \caption{UML graph for \lstinline|DrawableLoggingDeck|.}\label{fig:drawableloggingdeck}
\end{figure*}
If we try to keep the triangle inheritance, but apply dynamic dispatch, calling \lstinline|shuffleAndDraw| would trigger ambiguity. 
In principle we want \lstinline|shuffleAndDraw| to invoke \lstinline|LoggingDeck.draw| because they belong to the same branch, 
nonetheless, choosing either static dispatch
or dynamic dispatch cannot ensure both the triangle inheritance and code reuse. As far as we know, C++ does not have a solution
to it. Therefore we need to find another algorithm for method resolution.\\

@@

On the other hand, \lstinline|e.m()| where \lstinline|e| has static type \lstinline|I|, simply behaves the same as \lstinline|e.I::m()| in our model. Such a dispatch algorithm makes use of both the static type and the dynamic type of the receiver, so it is a combination of static and dynamic dispatch. Intuitively, the static type specifies one branch to avoid ambiguity, and the dynamic type finds the latest version on that branch. It may still introduce ambiguity when there are multiple paths from the static type to the dynamic type, and those paths cause conflicts. We disallow this kind of conflict to ensure unambiguity. That is to say, we do not allow two methods to override a same base method when they cause conflicts. This is natural, as they are just two versions of the same operation, hence it is no longer an ``unintentional'' conflict, but the diamond problem.
